---
title: "Exploring Shots Data"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

<style>
body{
    font-size: 14pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(here)
library(glue)
library(tidyverse)
library(Matrix)
library(hockeyR)


```


# Intro 

Terminology:

* "Shots on goal" = A shot that directs the puck towards the net and either goes into the net for a goal or is stopped by the goaltender for a save. In our data set, `n_sog` = `n_goal`+`n_shot`

* "Shot Attempts" = All shots, including shots on goal, missed shots and blocked shots. In our data set, `n_sa` =`n_goal`+`n_shot`+`n_missshot`+`n_blkdshot`. (See https://www.nhl.com/blackhawks/news/introducing-shot-attempts-sat/c-755520, this metric is tracked by the NHL)


Here are a few example rows: 
```{r}
data_fname = "data/sog-data_s'18'19'20'21_2022-04-01.rds"
shifts <- readRDS(here(data_fname))
shifts <- as.tibble(shifts) %>%
  mutate(n_sog_home=n_shot_home+n_goal_home, n_sog_away= n_shot_away+n_goal_away, n_sog=n_sog_home+n_sog_away) %>%
  mutate(n_sa_home = n_sog_home + n_blkdshot_home + n_missshot_home, n_sa_away = n_sog_away + n_blkdshot_away + n_missshot_away, n_sa=n_sa_away+n_sa_home)

shifts %>% select(shift_time, n_sog_home, n_sog_away, n_sog, n_sa_home, n_sa_away, n_sa)

```


The basic distribution of shift duration is as follows:
```{r}
ggplot(shifts) +
  geom_histogram(aes(x=shift_time), color="black", fill="lightblue", binwidth = 5, boundary=0) +
  scale_x_continuous(breaks = seq(0, 150, by = 10)) +
  theme_bw() +
  labs(x = "Shift Duration (s)")
  
```

Note that while they don't show up in this histogram, there are still a decent number of shifts that lasted for longer than a minute:

```{r, echo=TRUE}
sum(shifts$shift_time>60)
```


# Analysis of Shots on Goal

The vast majority of shifts had 0 shots on goal:
```{r}
shifts %>%
  ggplot(aes(x=n_sog)) +
  geom_bar(fill= "lightblue", color="black") +
  theme_bw() +
  labs(x="Number of shots on Goal")

```
Sort of a geometric distribution or perhaps extremely zero-inflated Poisson?

We're interested in the relationship between shift duration and number of shots on goals:
```{r}

shifts %>%
  ggplot(aes(x=shift_time, y=n_sog)) +
  geom_point(position = position_jitter(height=.2), alpha=.2) +
  scale_x_continuous(breaks = seq(0, 150, by = 15)) +
  theme_bw() +
  labs(x="Shift duration", y="Shots on goal")

```
```{r, echo=TRUE}
cor(shifts$shift_time, shifts$n_sog)
```

The maximum number of shots on goal in any shift was 7. While there is some positive correlation, there were many short shifts (<10 s) with several shots on goal and there were also many long shifts with no shots on goal. 


We were also interested in the relationship between the number of shots on goals by the two different teams on the ice:
```{r}
sog_counts <- as.data.frame(table(shifts$n_sog_home, shifts$n_sog_away)) %>%
  rename(n_sog_home = Var1, n_sog_away=Var2) %>%
  filter(Freq>0) %>%
  mutate(Freq = log(Freq))


ggplot(sog_counts, aes(n_sog_home, n_sog_away)) +
  geom_tile(aes(fill = Freq), colour = "black") +
  scale_fill_gradient(low = "white", high = "steelblue", name="Log of count") +
  geom_text(aes(label = round(exp(Freq)))) +
  labs(x="Home team's number of shots on goal", y= "Away team's number of shots on goal") +
  theme_minimal()

```
The counts are uncorrelated.


One thing we might do is divide shots on goals by shift duration to obtain a rate of shots on goal.
Here, we plot rate of shots on goal versus shift duration
```{r}

shifts <- shifts %>%
  mutate(rate_sog_home = n_sog_home/shift_time, rate_sog_away = n_sog_away/shift_time,
         rate_sog = rate_sog_home+rate_sog_away)


ggplot(shifts, aes(x=shift_time, y=rate_sog)) +
  geom_point(alpha=.3) + 
  theme_bw() +
  labs(x="Shift duration", y="Rate of shots on goal")



```
There's some negative correlation: shorter shifts have higher rates. The distinct curves are due to the discreteness of the shots on goal counts.  


# Analysis of Shot Attempts

The vast majority of shifts had 0 shot attempts:
```{r}
shifts %>%
  ggplot(aes(x=n_sa)) +
  geom_bar(fill= "lightblue", color="black") +
  theme_bw() +
  labs(x="Number of shots attempts")

```
Relationship between shift duration and number of shots on goals:
```{r}
shifts %>%
  ggplot(aes(x=shift_time, y=n_sa)) +
  geom_point(position = position_jitter(height=.2), alpha=.2) +
  scale_x_continuous(breaks = seq(0, 150, by = 15)) +
  theme_bw() +
  labs(x="Shift duration", y="Number of shot attempts")

```
```{r, echo=TRUE}
cor(shifts$shift_time, shifts$n_sa)
```
More positively correlated than shots on goal with shift duration. Still, plenty of long shifts with no shot attempts. 

Home versus away team shot attempts still largely uncorrelated:
```{r}
sog_counts <- as.data.frame(table(shifts$n_sa_home, shifts$n_sa_away)) %>%
  rename(n_sa_home = Var1, n_sa_away=Var2) %>%
  filter(Freq>0) %>%
  mutate(Freq =log(Freq))


ggplot(sog_counts, aes(n_sa_home, n_sa_away)) +
  geom_tile(aes(fill = Freq), colour = "black") +
  scale_fill_gradient(low = "white", high = "steelblue", name="Log of count") +
  geom_text(aes(label = exp(Freq)), size=3) +
  labs(x="Home team's number of shots attempts", y= "Away team's number of shot attempts") +
  theme_minimal()

```

Finally, we can divide by shift time and look at rates:
```{r}

shifts <- shifts %>%
  mutate(rate_sa_home = n_sa_home/shift_time, rate_sa_away = n_sa_away/shift_time,
         rate_sa = rate_sa_home+rate_sa_away)


ggplot(shifts, aes(x=shift_time, y=rate_sa)) +
  geom_point(alpha=.3) + 
  theme_bw() +
  labs(x="Shift duration", y="Rate of shot attempts")


```
As we now expect from having seen the same plot for SOG, shorter shifts also have higher rates of shot attempts.



# Relationship between shots on goal and blocked shots/missed shots


```{r}
sog_counts <- as.data.frame(table(shifts$n_sog, shifts$n_sa-shifts$n_sog)) %>%
  rename(n_sog = Var1, n_blckdandmissed=Var2) %>%
  filter(Freq>0) %>%
  mutate(Freq = log(Freq)) 


ggplot(sog_counts, aes(n_sog, n_blckdandmissed)) +
  geom_tile(aes(fill = Freq), colour = "black") +
  scale_fill_gradient(low = "white", high = "steelblue", name="Log of count") +
  geom_text(aes(label = round(exp(Freq),1))) +
  labs(x="Number of shots on goal", y= "Number of blocked and missed shots") +
  theme_minimal()

```


# How do these rates (of goals scored, shots on goal, and shot attempts) translate to league points at the team level?  

```{r}

# Team goal rates
team_as_home_goal <-shifts %>%
  group_by(home_abbreviation) %>% 
  summarise(n_goal_home = sum(n_goal_home), shift_time_home =sum(shift_time)) %>%
  rename(team = home_abbreviation)
  
team_as_away_goal <- shifts %>% 
  group_by(away_abbreviation) %>% 
  summarise(n_goal_away = sum(n_goal_away), shift_time_away=sum(shift_time)) %>%
  rename(team = away_abbreviation)

team_goal <- inner_join(team_as_home_goal, team_as_away_goal) %>%
  mutate(n_goal = n_goal_home + n_goal_away, rate_goal = n_goal/(shift_time_home+shift_time_away)) %>%
  select(team, n_goal, rate_goal)

#Team shots on goal rates
team_as_home_sog <-shifts %>%
  group_by(home_abbreviation) %>% 
  summarise(n_sog_home = sum(n_sog_home), shift_time_home =sum(shift_time)) %>%
  rename(team = home_abbreviation)
  
team_as_away_sog <- shifts %>% 
  group_by(away_abbreviation) %>% 
  summarise(n_sog_away = sum(n_sog_away), shift_time_away=sum(shift_time)) %>%
  rename(team = away_abbreviation)

team_sog <- inner_join(team_as_home_sog, team_as_away_sog) %>%
  mutate(n_sog = n_sog_home + n_sog_away, rate_sog = n_sog/(shift_time_home+shift_time_away)) %>%
  select(team, n_sog, rate_sog)

#Team shot attempts rates
team_as_home_sa <-shifts %>%
  group_by(home_abbreviation) %>% 
  summarise(n_sa_home = sum(n_sa_home), shift_time_home =sum(shift_time)) %>%
  rename(team = home_abbreviation)
  
team_as_away_sa <- shifts %>% 
  group_by(away_abbreviation) %>% 
  summarise(n_sa_away = sum(n_sa_away), shift_time_away=sum(shift_time)) %>%
  rename(team = away_abbreviation)

team_sa <- inner_join(team_as_home_sa, team_as_away_sa) %>%
  mutate(n_sa = (n_sa_home + n_sa_away), rate_sa = n_sa/(shift_time_home+shift_time_away)) %>%
  select(team, n_sa, rate_sa)


#Combine
teams <- inner_join(team_goal, team_sog) %>%
  inner_join(., team_sa) %>%
  filter(team != "VGK") #Weird numbers for this team
  

#Team records 
records <- rbind(get_team_records(2018), get_team_records(2019), get_team_records(2020), get_team_records(2021)) %>% 
  group_by(team_abbr) %>%
  summarise(points = sum(st_points)) %>% 
  rename(team = team_abbr)
  
  
#Join records with shot data 
  
teams <- inner_join(teams, records)



```


```{r}
library(gridExtra)
pgoals <- ggplot(teams, aes(x=rate_goal, y=points)) +
  geom_point() +
  theme_bw() +
  labs(x="Rate of Goals Scored", y="League Points") 
  
pgoals = pgoals + geom_smooth(method='lm', formula= y~x)

psog <- ggplot(teams, aes(x=rate_sog, y=points)) +
  geom_point() +
  theme_bw() +
  labs(x="Rate of Shots on Goal", y="League Points") 

psog = psog + geom_smooth(method='lm', formula= y~x)

psa <- ggplot(teams, aes(x=rate_sa, y=points)) +
  geom_point() +
  theme_bw() +
  labs(x="Rate of Shot Attempts", y="League Points") 

psa = psa + geom_smooth(method='lm', formula= y~x)


grid.arrange(pgoals, psog, psa)


```





