---
title: "Interpretation"
author: "Evan Gorstein"
date: "5/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    cache = FALSE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center"
)

library(tidyverse)
library(here)
library(glue)
library(rstan)
library(bayesplot)
library(patchwork)
library(plotly)
library(scales)
library(knitr)
library(kableExtra)
source(here("analysis/utils.R"))

bayesplot::color_scheme_set("red")

# Tables
kable_standard <- function(...) {
    kable(booktabs = TRUE, linesep = "", ...) %>% 
        kable_styling(full_width = FALSE)
}

options(knitr.kable.NA = '')

```



```{r}
#Pull in model
mod_dir <- "model/shots/output"
model_fname <- "ppool_sh-go_nt_s'21_2022-05-09.rds"
pm_fit <- readRDS(here(mod_dir,model_fname))

#Pull in data
data_dir <- "data"
data_fname <- "sog-model-data_o-sh-go_s'21_2022-04-25.rds"
model_data <- readRDS(here(data_dir, data_fname))


names = model_data@Dimnames[[2]]
team_names <- names[seq(3, 63, by=2)] %>%
  str_sub(start=1, end=3)

np = which(names == names[65])[2]-65 #Number of non-goalie players
ng = ncol(model_data) - 2*np - 64 #Number of total players
ng_player_names <- names[64+1:np] #Non-goalie player names 
player_names <-  names[64+np+1:(np+ng)]
model <- pull_modname(model_fname)
```

We interpret the effects in the shots-on-goal model without teams for the 2020-2021 season. 

Model: `r model_fname`  
Data: `r data_fname`

The sampling model is $y \sim \text{Pois}(\lambda)$, where 

\begin{align}
\lambda &= \exp\{\mu + \beta^{off}_{1} + \ldots + \beta^{off}_{5} - \beta^{def}_{1} - \ldots - \beta^{def}_{5} - \beta^{def}_{6} + \log(t)\} \\
&= \exp\{\beta^{off}_{1}\}*\exp\{\mu + \beta^{off}_{2} + \ldots + \beta^{off}_{5} - \beta^{def}_{1} - \ldots - \beta^{def}_{5} - \beta^{def}_{6} + \log(t)\}
\end{align}

We see that player 1 of the offense being on the ice produces a multiplicative effect of $\exp(\beta^{off}_{1})$ on the expected value of their team's shots on goal during the shift. Here is the posterior distribution for this effect for Alex Ovechkin

```{r}
alex_index = which(player_names == "Alex.Ovechkin") %>% as.character()
alex_draws = extract(pm_fit, pars = paste0("beta_off[", alex_index, "]")) %>% unlist()
mult_effect = exp(alex_draws)
hist(mult_effect, main = "Posterior distribution of multiplicative effect\n for Alex Ovechkin")
```

The player with the largest posterior mean effect was Connor McDavid.  

```{r}
connor_index = which(player_names == "Connor.McDavid") %>%
  as.character()
connor_draws = extract(pm_fit, pars = paste0("beta_off[", connor_index, "]")) %>%
  unlist()
mult_effect = exp(connor_draws)
hist(mult_effect, main = "Posterior distribution of multiplicative effect\n for Connor McDavid")
```

Let's suppose the Oilers are playing forwards Jesse Puljujarvi, Ryan Nugent-Hopkins and defensemen Tyson Barrie and Darnell Nurse (see [historical line-up combinations](https://frozenpool.dobbersports.com/frozenpool_linecombo.php?select=TF&games=2020-2021%3AR%3A99&period=ALL&situation=EV)) against a team with completely average defense. If we consider a shift lasting 13s, the mean in our data set, we can find the posterior distribution of the expected number of shots on goal for the Oilers during a shift with Connor McDavid included in the line-up versus a shift with McDavid replaced by a completely average offensive player.

```{r}
mu_draws <- extract(pm_fit, pars ="mu") %>% unlist()
oiler_players = c("Jesse.Puljujarvi" , "Ryan.Nugent.Hopkins", "Tyson.Barrie", "Darnell.Nurse")
oiler_indices = which(player_names %in% oiler_players) %>% 
  as.character()
oiler_pars = paste0("beta_off[", oiler_indices, "]")
draws = extract(pm_fit, pars = oiler_pars) %>% 
  data.frame() %>%
  mutate(mu_draws = mu_draws, logt = log(13), connor_draws=connor_draws) %>%
  mutate(wo_connor = exp(mu_draws + beta_off.24. + beta_off.31. + 
           beta_off.634. + beta_off.813. + logt), 
         wt_connor = wo_connor*exp(connor_draws))
  
ggplot(draws) +
  labs(x = "Expected shots on goal", fill="Line-up") +
  geom_histogram(aes(x = wo_connor, fill = "With Average Player"), alpha =.2, color="black") +
  geom_histogram(aes(x = wt_connor, fill = "With McDavid"), alpha =.2, color= "black") +
  scale_fill_manual(values = c("red","blue")) 

```

#Offense

```{r}
mu_draws <- extract(pm_fit, pars = "mu") %>% unlist()

offense_draws <- extract(pm_fit, pars ="beta_off") %>%
  data.frame() %>%
  setNames(ng_player_names) %>%
  rowwise() %>%
  mutate(m = median(c_across(everything()))) %>%
  ungroup()

defense_draws <- extract(pm_fit, pars ="beta_def") %>%
  data.frame() %>%
  setNames(player_names)

shift_times = model_data[, 2]

nr = nrow(model_data)
nc = ncol(model_data)

shifts = as.matrix(model_data[, 65:nc])
indices = which(shifts !=0, arr.ind =T) %>%
  data.frame() %>%
  rename(shift = row, player_index = col) %>%
  arrange(shift)


## Given a shift number, for each player on the ice for the offensive team for this shift, this function returns the distribution of their contribution to their team's shots on goal relative to the median player. So the returned object is a 4,000 by 5 matrix. Actually, to facilitate the calculation of the running total across all the shifts in the season in the for loop that follows, it will return a list whose first element is the aforementioned matrix and whose second entry is a vector of the indices of these players. 
form_post <- function(shift_no) {
  players = indices %>% 
    filter(shift == shift_no)
  
  offense = offense_draws %>% 
    select(players$player_index[1:5]) %>%
    rowSums()
  defense = defense_draws %>% 
    select(players$player_index[6:11]-np) %>%
    rowSums()
  
  Z = mu_draws + offense - defense + log(shift_times[shift_no])      
  m = matrix(0, nrow = 4000, ncol = 5)
  
  for (i in 1:5) {
    Z_med = Z - offense_draws[[players$player_index[i]]] + offense_draws[[np+1]] 
    m[,i] = exp(Z) - exp(Z_med)
  }
  
  return(list(m, players$player_index[1:5]))
}

season_cont = matrix(0, nrow = 4000, ncol = np)

for (j in 1:nr){
  shift_cont = form_post(j)
  season_cont[, shift_cont[[2]]] = season_cont[, shift_cont[[2]]] + shift_cont[[1]]
  print(j)
} 

matrix_fname <- glue::glue("shgo_nt_off_21_{lubridate::today()}.rds")
saveRDS(season_cont, here("analysis", "output", matrix_fname))

```


```{r}
hist(season_cont[,333]) #ovechkin
hist(season_cont[,593]) #mcdavid
colMeans(season_cont)
hist(season_cont[,587]) #Patrice Bergerson
hist(season_cont[,588]) #Adam Larsson
```








