---
title: "What's our model learning?"
author: "Evan Gorstein"
date: "5/9/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    cache = FALSE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center"
)

library(tidyverse)
library(here)
library(glue)
library(rstan)
library(bayesplot)
library(patchwork)
library(plotly)
library(scales)
library(knitr)
library(kableExtra)
source(here("analysis/utils.R"))

bayesplot::color_scheme_set("red")

# Tables
kable_standard <- function(...) {
    kable(booktabs = TRUE, linesep = "", ...) %>% 
        kable_styling(full_width = FALSE)
}

options(knitr.kable.NA = '')

```



```{r}
#Pull in model
mod_dir <- "model/shots/output"
model_fname <- "ppool_sh-go_nt_s'21_2022-05-09.rds"
pm_fit <- readRDS(here(mod_dir,model_fname))

#Pull in data
data_dir <- "data"
data_fname <- "sog-model-data_o-sh-go_s'21_2022-04-25.rds"
model_data <- readRDS(here(data_dir, data_fname))


names = model_data@Dimnames[[2]]
team_names <- names[seq(3, 63, by=2)] %>%
  str_sub(start=1, end=3)

np = which(names == names[65])[2]-65 #Number of non-goalie players
ng = ncol(model_data) - 2*np - 64 #Number of total players
ng_player_names <- names[64+1:np] #Non-goalie player names 
player_names <-  names[64+np+1:(np+ng)]
model <- pull_modname(model_fname)
```

Model: `r model_fname`
Data: `r data_fname`

```{r}
fit_summary <- summary(pm_fit) #Posterior summary with row for each parameter

data_cols <- list(
  "off" = c(1, 2, 64 + 1:np),
  "def" = c(1, 2, 64 + np + 1:(np+ng))
)

fit_rows <- list(
  "off" = 1 + (1:np),
  "def" = (2+np):(2*np+ng+1)
)

# Totals for player's rows
sum_player_var = function(player, var, data_set) {
  shifts = which(data_set[, player] != 0)
  sum(data_set[shifts, var])
}
sum_not_player_var = function(player, var, data_set) {
  shifts = which(data_set[, player] == 0)
  sum(data_set[shifts, var])
}

get_player_summary <- function(df, fit, data_cols, fit_rows, players) {
  
  model_data_sm <- df[, data_cols]
  out <- data.frame(fit$summary[fit_rows,])
  row.names(out) <- players

  out$appear <- abs(Matrix::colSums(model_data_sm)[-c(1,2)])
  
  out <- out %>% 
    mutate(
      ice_time = map_dbl(players, ~sum_player_var(.x, "shift_time", model_data_sm)),
      sog = map_dbl(players, ~sum_player_var(.x, "y", model_data_sm)),
      not_ice_time = map_dbl(players, ~sum_not_player_var(.x, "shift_time", model_data_sm)),
      not_sog = map_dbl(players, ~sum_not_player_var(.x, "y", model_data_sm))
    )

}

off_players <- get_player_summary(model_data, fit_summary, 
                                  data_cols[["off"]], fit_rows[["off"]], ng_player_names)

def_players <- get_player_summary(model_data, fit_summary, 
                                  data_cols[["def"]], fit_rows[["def"]], player_names)


def_players_ng = def_players[1:np,] #non-goalie defensive effects
def_players_g = def_players[(np+1):nrow(def_players),]

```


## Checking relationship between number of appearances in data and effect estimates

```{r}

plot_appear <- function(df, title) {
  ggplot(df, aes(x=appear, y=X50.)) + 
    geom_point() +
    labs(
      x = "Number of Shifts",
      y = "Posterior Median",
      title = title 
    ) + 
    theme_minimal()
}

plot_appear(off_players, 
            "Offensive effect posterior median \nversus number of shifts for all players")

plot_appear(def_players_ng, 
  "Defensive effect posterior median \nversus number of shifts for non-goalies")

plot_appear(def_players_g, 
            "Defensive effect posterior median \nversus number of shifts for goalies")

```


## Checking relationship between number of appearances in data and length of credible intervals 

```{r}
plot_shift_vs_length <- function(df, title) {
  df %>% 
    mutate(length_50 = X75. - X25.) %>%
    ggplot(aes(x=appear, y=length_50)) +
    geom_point() +
    labs(
      x = "Number of shifts",
      y = "Length of 50% credible interval",
      title = title
    )
}

plot_shift_vs_length(off_players, "Offensive effects, all players")
plot_shift_vs_length(def_players_ng, "Defensive effects, non-goalies")
plot_shift_vs_length(def_players_g, "Defensive effects, goalies")
```

## Checking relationship between number of shots on goal while on ice and effect estimates

```{r}
plot_outcome_vs_pmean <- function(df, title) {
  df %>%
    rownames_to_column() %>% 
    mutate(sog_rate = sog/ice_time) %>%
    mutate(text = glue(
      "{rowname} 
      posterior median: {number(X50., 0.001)}
      SOG rate: {number(sog_rate, 0.0001)}
      n_appear: {appear}"
    )) %>% 
    ggplot(aes(x=sog_rate, y=X50., text = text)) + 
    geom_point() +
    theme_minimal() +
    labs(
      x = "Shots on goal per second",
      y = "Effect posterior median",
      title = title
    )
}

plot_outcome_vs_pmean(off_players, "Offensive effect estimate \nversus teams' shots on goal rate while player is on ice") %>% ggplotly(tooltip = "text")
plot_outcome_vs_pmean(def_players_ng, "Defensive effect estimate versus opposing \nteams' shots on goal rate while player is on ice") %>% ggplotly(tooltip = "text")
plot_outcome_vs_pmean(def_players_g, "Defensive effect posterior median \nversus opposing teams' shots on goal per second while player is on ice") %>% ggplotly(tooltip = "text")


```


```{r}
plot_outcome_diff_vs_pmean <- function(df, title) {
  df %>%
    rownames_to_column() %>% 
    mutate(sog_rate_diff = sog/ice_time - not_sog/not_ice_time) %>%
    mutate(text = glue(
      "{rowname} 
      posterior median: {number(X50., 0.001)}
      SOG rate: {number(sog/ice_time, 0.0001)}
      SOG rate diff: {number(sog_rate_diff, 0.0001)}
      n_appear: {appear}"
    )) %>% 
    ggplot(aes(x=sog_rate_diff, y=X50., text = text)) + 
    geom_point() +
    theme_minimal() +
    labs(
      x = "Difference in shots on goal per second when on vs off ice",
      y = "Effect posterior median",
      title = title
    )
}

plot_outcome_diff_vs_pmean(off_players, "Offensive effect estimate \nversus shots on goal rate difference while player is on ice vs not") %>% ggplotly(tooltip = "text")
plot_outcome_diff_vs_pmean(def_players_ng, "Defensive effect estimate versus opposing \nversus shots on goal rate difference while player is on ice vs not") %>% ggplotly(tooltip = "text")
plot_outcome_diff_vs_pmean(def_players_g, "Defensive effect posterior median \nversus shots on goal rate difference while player is on ice vs not") %>% ggplotly(tooltip = "text")


```


## Investigate players with high discrepancy from shots on goal per second

```{r}

table_top_20_resid <- function(df, title) {
  
  fit <- lm(X50. ~ sog_rate, data = df %>% mutate(sog_rate = sog/ice_time))
  df$resid <- residuals(fit)
  
  df %>% 
    mutate(sog_rate = sog/ice_time) %>% 
    arrange(desc(abs(resid))) %>% 
    rownames_to_column() %>% 
    slice_head(n = 20) %>% 
    select(rowname, mean, X50., sog_rate, resid, appear) %>%
    kable_standard(caption = title,
                   digits = 3,
                   col.names = c("Player", "Mean", "Median", "SOG rate", "Residual", "N_appearances"))
}

table_top_20_resid(off_players, title = "Top 20 effects by residual, offensive")
table_top_20_resid(def_players_ng, title = "Top 20 effects by residual, defensive non-goalie")
table_top_20_resid(def_players_g, title = "Top 20 effects by residual, defensive goalie")

```



## Check for players that are uni-dimensional, high in only offense or defense

```{r}


all_players <- 
  inner_join(
    off_players %>% rownames_to_column(),
    def_players_ng %>% rownames_to_column(),
    by = "rowname",   
    suffix = c("_off", "_def")
  )


p_off_def <- all_players %>% 
  mutate(text = glue(
      "{rowname} 
      Offensive effect: {number(X50._off, 0.001)}
      Defensive effect: {number(X50._def, 0.001)}
      n_appear: {appear_off}"
    )) %>% 
  ggplot(aes(x = X50._off, y = X50._def, text = text)) +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_vline(xintercept = 0, color = "grey50") + 
  geom_point() +
  labs(
    x = "Offensive effect, posterior median",
    y = "Defensive effect, posterior median",
    title = "Comparison of offensive and defensive effects"
  )
  

p_off_def %>% 
  ggplotly(tooltip = "text")
  

```







