---
title: "Interpretation"
author: "Evan Gorstein"
date: "5/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    cache = FALSE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center"
)

library(tidyverse)
library(here)
library(glue)
library(rstan)
library(bayesplot)
library(patchwork)
library(plotly)
library(scales)
library(knitr)
library(kableExtra)
source(here("analysis/utils.R"))

bayesplot::color_scheme_set("red")

# Tables
kable_standard <- function(...) {
    kable(booktabs = TRUE, linesep = "", ...) %>% 
        kable_styling(full_width = FALSE)
}

options(knitr.kable.NA = '')

```



```{r}
#Pull in model
mod_dir <- "model/shots/output"
model_fname <- "ppool_sh-go_nt_s'21_2022-05-09.rds"
pm_fit <- readRDS(here(mod_dir,model_fname))

#Pull in data
data_dir <- "data"
data_fname <- "sog-model-data_o-sh-go_s'21_2022-04-25.rds"
model_data <- readRDS(here(data_dir, data_fname))


names = model_data@Dimnames[[2]]
team_names <- names[seq(3, 63, by=2)] %>%
  str_sub(start=1, end=3)

np = which(names == names[65])[2]-65 #Number of non-goalie players
ng = ncol(model_data) - 2*np - 64 #Number of total players
ng_player_names <- names[64+1:np] #Non-goalie player names 
player_names <-  names[64+np+1:(np+ng)]
model <- pull_modname(model_fname)
```

We interpret the effects in the shots-on-goal model without teams for the 2020-2021 season. 

Model: `r model_fname`  
Data: `r data_fname`

The sampling model is $y \sim \text{Pois}(\lambda)$, where 

\begin{align}
\lambda &= \exp\{\mu + \beta^{off}_{1} + \ldots + \beta^{off}_{5} - \beta^{def}_{1} - \ldots - \beta^{def}_{5} - \beta^{def}_{6} + \log(t)\} \\
&= \exp\{\beta^{off}_{1}\}*\exp\{\mu + \beta^{off}_{2} + \ldots + \beta^{off}_{5} - \beta^{def}_{1} - \ldots - \beta^{def}_{5} - \beta^{def}_{6} + \log(t)\}
\end{align}

We see that player 1 of the offense being on the ice produces a multiplicative effect of $\exp(\beta^{off}_{1})$ on the expected value of their team's shots on goal during the shift. Here is the posterior distribution for this effect for Alex Ovechkin

```{r}
alex_index = which(player_names == "Alex.Ovechkin") %>% as.character()
alex_draws = extract(pm_fit, pars = paste0("beta_off[", alex_index, "]")) %>% unlist()
mult_effect = exp(alex_draws)
hist(mult_effect, main = "Posterior distribution of multiplicative effect\n for Alex Ovechkin")
```

The player with the largest posterior mean effect was Connor McDavid.  

```{r}
connor_index = which(player_names == "Connor.McDavid") %>%
  as.character()
connor_draws = extract(pm_fit, pars = paste0("beta_off[", connor_index, "]")) %>%
  unlist()
mult_effect = exp(connor_draws)
hist(mult_effect, main = "Posterior distribution of multiplicative effect\n for Connor McDavid")
```

Let's suppose the Oilers are playing forwards Jesse Puljujarvi, Ryan Nugent-Hopkins and defensemen Tyson Barrie and Darnell Nurse (see [historical line-up combinations](https://frozenpool.dobbersports.com/frozenpool_linecombo.php?select=TF&games=2020-2021%3AR%3A99&period=ALL&situation=EV)) against a team with completely average defense. If we consider a shift lasting 13s, the mean in our data set, we can find the posterior distribution of the expected number of shots on goal for the Oilers during a shift with Connor McDavid included in the line-up versus a shift with McDavid replaced by a completely average offensive player.

```{r}
mu_draws <- extract(pm_fit, pars ="mu") %>% unlist()
oiler_players = c("Jesse.Puljujarvi" , "Ryan.Nugent.Hopkins", "Tyson.Barrie", "Darnell.Nurse")
oiler_indices = which(player_names %in% oiler_players) %>% 
  as.character()
oiler_pars = paste0("beta_off[", oiler_indices, "]")
draws = extract(pm_fit, pars = oiler_pars) %>% 
  data.frame() %>%
  mutate(mu_draws = mu_draws, logt = log(13), connor_draws=connor_draws) %>%
  mutate(wo_connor = exp(mu_draws + beta_off.24. + beta_off.31. + 
           beta_off.634. + beta_off.813. + logt), 
         wt_connor = wo_connor*exp(connor_draws))
  
ggplot(draws) +
  labs(x = "Expected shots on goal", fill="Line-up") +
  geom_histogram(aes(x = wo_connor, fill = "With Average Player"), alpha =.2, color="black") +
  geom_histogram(aes(x = wt_connor, fill = "With McDavid"), alpha =.2, color= "black") +
  scale_fill_manual(values = c("red","blue")) 

```








